---
title: "Evaluating a Bayesian GSD: conjugate and non-conjugate priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evaluating a Bayesian GSD: conjugate and non-conjugate priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
set.seed(1L)
```

This vignette walks through the **evaluation** workflow of a Bayesian
group sequential design: given a fixed design, prior and decision rule,
estimate the operating characteristics by Monte Carlo simulation under a
user-supplied data-generating effect. We work through two scenarios on the
same binary ADRENAL re-design, differing only in the prior:

1. **Conjugate prior** --- independent $\mathrm{Beta}(1, 1)$ priors on the
   per-arm response rates. Each per-arm posterior stays a single beta, and
   the risk-difference tail probability, which has no closed form, is
   evaluated by a fixed 64-node Gauss--Legendre quadrature at every look
   (no MCMC).
2. **Non-conjugate prior** --- a logit-normal prior on the control-arm
   rate, approximated by a beta mixture via `fit_mixture()`. The same
   quadrature applies component-wise, so the per-look tail-probability
   cost scales with the number of mixture components.

The shared design and decision rule are constructed once.

```{r design, message = FALSE}
library(adabay)

des <- set_design(
  endpoint     = "binary",
  n_per_look   = c(760, 1520, 2280, 3040, 3800),
  effect_scale = "risk_difference",
  alternative  = "less"
)

dec <- set_decision(
  efficacy = list(list(threshold_effect = 0, threshold_prob = 0.99)),
  futility = list(list(threshold_effect = 0, threshold_prob = 0.90)),
  futility_binding = FALSE
)
```

Because `futility_binding = FALSE`, the futility boundary is **non-binding**:
the reported Type I error rate (`$alpha`) is computed *as if the futility rule
were ignored*, which is the regulatory convention, whereas the realised power
under the actual trial conduct --- in which the trial does stop at a futility
crossing --- is `$eff_prob`. The two coincide only under binding futility. We
therefore read Type I error from `$alpha` and power from `$eff_prob` below.

# 1. Conjugate prior

The simplest case: independent $\mathrm{Beta}(1, 1)$ priors on the per-arm
response rates.

```{r conj-prior}
pri_conj <- set_prior(
  endpoint = "binary",
  arms = list(c = list(family = "beta", a = 1, b = 1),
              t = list(family = "beta", a = 1, b = 1))
)
pri_conj
```

We evaluate the operating characteristics under both the null
($\theta_c = \theta_t = 0.33$) and the alternative ($\theta_c = 0.33$,
$\theta_t = 0.28$), with `n_trials = 2000` for a fast vignette. Raising
this to `n_trials = 1e6` recovers the manuscript precision.

```{r conj-eval, message = FALSE}
oc_conj_h0 <- evaluate_design(
  des, pri_conj, dec,
  effect   = list(theta_c = 0.33, theta_t = 0.33),
  n_trials = 2000, cores = 1, seed = 1L
)
oc_conj_h1 <- evaluate_design(
  des, pri_conj, dec,
  effect   = list(theta_c = 0.33, theta_t = 0.28),
  n_trials = 2000, cores = 1, seed = 1L
)

cat(sprintf("Conjugate Beta(1,1) prior:\n"))
cat(sprintf("  Type I error rate  = %.2f%%\n", 100 * oc_conj_h0$alpha))
cat(sprintf("  Power              = %.2f%%\n", 100 * oc_conj_h1$eff_prob))
cat(sprintf("  E(N | H1)          = %.0f patients\n",
            oc_conj_h1$expected_sample_size))
```

The per-look stopping probabilities and the bar chart show where the trial
accumulates efficacy and futility mass.

```{r conj-plot, fig.alt = "Per-look efficacy and futility stopping probabilities under H1 for the conjugate-prior design."}
if (requireNamespace("ggplot2", quietly = TRUE)) plot(oc_conj_h1)
```

# 2. Non-conjugate prior approximated by a beta mixture

Suppose the prior on the control-arm rate $\theta_c$ is summarised by a
logit-normal with logit-scale mean $-0.7$ and standard deviation $0.4$,
corresponding to a prior median near $0.33$ and modest spread. A
logit-normal is not conjugate to the binomial, so we approximate it by a
finite mixture of beta components via `fit_mixture()`. The treatment-arm
prior remains $\mathrm{Beta}(1, 1)$.

Because `prior` is supplied here as a density function rather than a
vector of pre-drawn samples, `fit_mixture()` draws `n_samples` observations
from it internally; passing `seed` makes that draw (and hence the fitted
mixture) reproducible regardless of what randomness this session has
already consumed, rather than depending on the ambient RNG state.

```{r nonconj-prior, message = FALSE, warning = FALSE}
logit_normal_density <- function(theta) {
  meanlog <- -0.7
  sdlog   <-  0.4
  z <- log(theta / (1 - theta))
  dnorm(z, meanlog, sdlog) / (theta * (1 - theta))
}

pri_c_arm <- fit_mixture(
  endpoint         = "binary",
  arm              = "c",
  prior            = logit_normal_density,
  n_components_max = 5L,
  tol_kl           = 1e-3,
  tol_tail         = 5e-3,
  n_samples        = 10000L,
  seed             = 1L
)
pri_c_arm

pri_nc <- set_prior(
  endpoint = "binary",
  arms     = list(c = pri_c_arm,
                  t = list(family = "beta", a = 1, b = 1))
)
```

`fit_mixture()` reports per-quantile tail-probability errors and emits a
runtime warning when the maximum absolute error exceeds `tol_tail`. The
overlay plot below compares the empirical-sample density against the
fitted mixture density, with vertical rugs at the tail-quantile grid.

```{r nonconj-overlay, fig.alt = "Empirical logit-normal prior density overlaid on the fitted beta-mixture density, with tail-quantile rugs."}
if (requireNamespace("ggplot2", quietly = TRUE)) plot(pri_c_arm)
```

The downstream evaluation is identical to the conjugate case --- there is
no change in user-facing code beyond passing `pri_nc` instead of
`pri_conj`.

```{r nonconj-eval, message = FALSE}
oc_nc_h0 <- evaluate_design(
  des, pri_nc, dec,
  effect   = list(theta_c = 0.33, theta_t = 0.33),
  n_trials = 2000, cores = 1, seed = 1L
)
oc_nc_h1 <- evaluate_design(
  des, pri_nc, dec,
  effect   = list(theta_c = 0.33, theta_t = 0.28),
  n_trials = 2000, cores = 1, seed = 1L
)

cat(sprintf("Non-conjugate (logit-normal) prior via beta mixture:\n"))
cat(sprintf("  Type I error rate  = %.2f%%\n", 100 * oc_nc_h0$alpha))
cat(sprintf("  Power              = %.2f%%\n", 100 * oc_nc_h1$eff_prob))
cat(sprintf("  E(N | H1)          = %.0f patients\n",
            oc_nc_h1$expected_sample_size))
```

# 3. Side-by-side comparison

The mixture prior is more informative than the flat $\mathrm{Beta}(1, 1)$
in the central region but matches it on the tails, so the operating
characteristics are close but not identical at the same simulation budget.

```{r compare}
data.frame(
  prior     = c("Beta(1,1) flat", "logit-normal mixture"),
  type_I_pp = c(100 * oc_conj_h0$alpha, 100 * oc_nc_h0$alpha),
  power_pp  = c(100 * oc_conj_h1$eff_prob, 100 * oc_nc_h1$eff_prob),
  E_N_H1    = c(oc_conj_h1$expected_sample_size,
                oc_nc_h1$expected_sample_size)
)
```

# Summary

`evaluate_design()` is the same call for the conjugate and the
non-conjugate case; only the `adabay_prior` object differs. The mixture
approximation enters the pipeline exactly once, when constructing the
`adabay_prior_arm` object via `fit_mixture()`, after which every downstream
entry point runs at full conjugate speed.
