---
title: "Calibrating a Bayesian GSD: conjugate and non-conjugate priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrating a Bayesian GSD: conjugate and non-conjugate priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

This vignette walks through the **calibration** workflow of a Bayesian
group sequential design: given a fixed design, prior, and operating-
characteristic targets, search a grid of candidate decision rules and
return the rule that meets the targets at minimum expected sample size.

The package decouples this search from the dominating cost --- the
trial-simulation pass --- via the caching API: `build_cache()` runs the
simulation pass once per scenario, and `calibrate_design()` or
`evaluate_design()` then scores any candidate threshold combination from
the cache in milliseconds.

We walk through the workflow on the ADRENAL re-design under two scenarios
that differ only in the prior:

1. **Conjugate prior** --- independent $\mathrm{Beta}(1, 1)$ priors.
2. **Non-conjugate prior** --- logit-normal on the control rate,
   approximated by a beta mixture via `fit_mixture()`.

The simulation budget is `n_trials = 5000` per cache to keep the vignette
quick. Raising to `n_trials = 1e5` (or more) recovers manuscript precision
with no change to the workflow.

```{r design, message = FALSE}
library(adabay)

des <- set_design(
  endpoint     = "binary",
  n_per_look   = c(760, 1520, 2280, 3040, 3800),
  effect_scale = "risk_difference",
  alternative  = "less"
)

R_CACHE <- 5000L

threshold_grid <- list(efficacy = 0, futility = 0)
```

# 1. Calibration under a conjugate prior

We build one cache under the null ($\theta_c = \theta_t = 0.33$) for the
type I error rate, and one under the alternative ($\theta_t = 0.28$) for
the power and the expected sample size. Both caches use the same fixed
seed, so every candidate threshold combination is scored on the same
simulated data.

```{r conj-prior, message = FALSE}
pri_conj <- set_prior(
  endpoint = "binary",
  arms = list(c = list(family = "beta", a = 1, b = 1),
              t = list(family = "beta", a = 1, b = 1))
)

cache_conj_h0 <- build_cache(
  des, pri_conj,
  effect   = list(theta_c = 0.33, theta_t = 0.33),
  n_trials = R_CACHE, cores = 1, seed = 1L,
  threshold_grid = threshold_grid
)
cache_conj_h1 <- build_cache(
  des, pri_conj,
  effect   = list(theta_c = 0.33, theta_t = 0.28),
  n_trials = R_CACHE, cores = 1, seed = 1L,
  threshold_grid = threshold_grid
)
```

## 1a. Common-across-looks calibration

`calibrate_design()` sweeps a grid of common-across-looks
posterior-probability thresholds $(p, q)$ against the type I error rate
upper bound and the power lower bound. Try the conventional
`power_target = 0.90` first:

```{r conj-calibrate-v1}
cal_conj_90 <- calibrate_design(
  cache         = cache_conj_h0,
  cache_alt     = cache_conj_h1,
  alpha_target  = 0.025,
  power_target  = 0.90,
  efficacy_grid = seq(0.95, 0.999, by = 0.005),
  futility_grid = seq(0.80, 0.95,  by = 0.05)
)
is.null(cal_conj_90$best)
```

No $(p, q)$ on this grid reaches 90% power at $\alpha \le 0.025$
(`calibrate_design()` warns and returns `best = NULL` rather than
silently picking a design that doesn't meet the targets). This is a
real feature of a *single, constant* threshold applied at every look: raising
$p$ to control type I error costs power at every look equally, including
the final one, so a **common** threshold cannot reach as high a power at
a given type I error as a threshold shaped across looks can (see 1b
below). Following the warning's own suggestion, relax the power target to
0.80:

```{r conj-calibrate}
cal_conj <- calibrate_design(
  cache         = cache_conj_h0,
  cache_alt     = cache_conj_h1,
  alpha_target  = 0.025,
  power_target  = 0.80,
  efficacy_grid = seq(0.95, 0.999, by = 0.005),
  futility_grid = seq(0.80, 0.95,  by = 0.05)
)
head(cal_conj$grid)
```

The grid records the type I error rate, the power and the expected
sample size at every $(p, q)$ combination. The best design at the
targets is `cal_conj$best` --- an `adabay_calibration_best` object
carrying `p`, `q`, `type_I` (from the $H_0$ cache), `power` and
`expected_sample_size` (from the $H_1$ cache) as plain fields, plus the
full `adabay_oc` objects for each hypothesis as `$oc_h0` / `$oc_h1`.

```{r conj-best}
if (!is.null(cal_conj$best)) {
  cal_conj$best
} else {
  cat("No (p, q) in the searched grid met both alpha and power targets",
      "at R =", R_CACHE, "per cache; raise R or widen the grid.\n")
}
```

## 1b. Stage-specific calibration

For Haybittle-Peto-style schedules (stringent interim thresholds, near-
nominal final), a short `evaluate_design()` loop against the same caches
is sufficient. No new simulation pass is required.

```{r conj-stage}
p_stage <- c(0.998, 0.998, 0.998, 0.998, 0.9775)
q_stage <- 0.90

dec_stage <- set_decision(
  efficacy = lapply(p_stage, function(p)
    list(list(threshold_effect = 0, threshold_prob = p))),
  futility = list(list(threshold_effect = 0, threshold_prob = q_stage)),
  futility_binding = TRUE
)

oc_conj_stage_h0 <- evaluate_design(cache_conj_h0, dec_stage)
oc_conj_stage_h1 <- evaluate_design(cache_conj_h1, dec_stage)

cat(sprintf("Conjugate prior, stage-specific p_1:5 = (0.998, 0.998, 0.998, 0.998, 0.9775):\n"))
cat(sprintf("  Type I error rate  = %.2f%%\n", 100 * oc_conj_stage_h0$alpha))
cat(sprintf("  Power              = %.2f%%\n", 100 * oc_conj_stage_h1$alpha))
cat(sprintf("  E(N | H1)          = %.0f patients\n",
            oc_conj_stage_h1$expected_sample_size))
```

The expected sample size and per-look stopping probabilities can be
inspected via the `print()` and `plot()` methods on the returned
`adabay_oc`.

```{r conj-plot, fig.alt = "Per-look stopping probabilities for the stage-specific calibrated design under the conjugate prior."}
if (requireNamespace("ggplot2", quietly = TRUE)) plot(oc_conj_stage_h1)
```

# 2. Calibration under a non-conjugate prior

The same workflow applies once we replace `pri_conj` with a mixture-
approximated logit-normal prior on the control rate.

```{r nonconj-prior, message = FALSE, warning = FALSE}
logit_normal_density <- function(theta) {
  meanlog <- -0.7
  sdlog   <-  0.4
  z <- log(theta / (1 - theta))
  dnorm(z, meanlog, sdlog) / (theta * (1 - theta))
}

pri_c_arm <- fit_mixture(
  endpoint         = "binary",
  arm              = "c",
  prior            = logit_normal_density,
  n_components_max = 5L,
  tol_kl           = 1e-3,
  tol_tail         = 5e-3,
  n_samples        = 10000L,
  seed             = 1L
)
pri_c_arm

pri_nc <- set_prior(
  endpoint = "binary",
  arms     = list(c = pri_c_arm,
                  t = list(family = "beta", a = 1, b = 1))
)
```

The two caches are built exactly as before, only the prior is different.

```{r nonconj-caches, message = FALSE}
cache_nc_h0 <- build_cache(
  des, pri_nc,
  effect   = list(theta_c = 0.33, theta_t = 0.33),
  n_trials = R_CACHE, cores = 1, seed = 1L,
  threshold_grid = threshold_grid
)
cache_nc_h1 <- build_cache(
  des, pri_nc,
  effect   = list(theta_c = 0.33, theta_t = 0.28),
  n_trials = R_CACHE, cores = 1, seed = 1L,
  threshold_grid = threshold_grid
)
```

## 2a. Common-across-looks calibration

Same `power_target = 0.80` as 1a, for the same reason: a single common
threshold across 5 looks cannot reach as high a power at this type I
error as a properly shaped (stage-specific) schedule can.

```{r nonconj-calibrate}
cal_nc <- calibrate_design(
  cache         = cache_nc_h0,
  cache_alt     = cache_nc_h1,
  alpha_target  = 0.025,
  power_target  = 0.80,
  efficacy_grid = seq(0.95, 0.999, by = 0.005),
  futility_grid = seq(0.80, 0.95,  by = 0.05)
)
head(cal_nc$grid)
cal_nc$best
```

## 2b. Stage-specific calibration

```{r nonconj-stage}
oc_nc_stage_h0 <- evaluate_design(cache_nc_h0, dec_stage)
oc_nc_stage_h1 <- evaluate_design(cache_nc_h1, dec_stage)

cat(sprintf("Non-conjugate prior, stage-specific p_1:5 = (0.998, 0.998, 0.998, 0.998, 0.9775):\n"))
cat(sprintf("  Type I error rate  = %.2f%%\n", 100 * oc_nc_stage_h0$alpha))
cat(sprintf("  Power              = %.2f%%\n", 100 * oc_nc_stage_h1$alpha))
cat(sprintf("  E(N | H1)          = %.0f patients\n",
            oc_nc_stage_h1$expected_sample_size))
```

# 3. Side-by-side calibrated designs

The mixture prior is more informative than the flat $\mathrm{Beta}(1, 1)$
in the central region, so for the same stage-specific schedule the type
I error rate and the power can shift modestly relative to the conjugate
case.

```{r compare-stage}
data.frame(
  prior     = c("Beta(1,1) flat", "logit-normal mixture"),
  type_I_pp = c(100 * oc_conj_stage_h0$alpha, 100 * oc_nc_stage_h0$alpha),
  power_pp  = c(100 * oc_conj_stage_h1$alpha, 100 * oc_nc_stage_h1$alpha),
  E_N_H1    = c(oc_conj_stage_h1$expected_sample_size,
                oc_nc_stage_h1$expected_sample_size)
)
```

# Summary

`build_cache()` runs the simulation pass exactly once per scenario.
Whatever the prior --- conjugate or a mixture-approximated non-conjugate
--- the calibration step is downstream of the cache and runs in
milliseconds per candidate threshold combination. The workflow is the
same in both prior scenarios; the only difference is the `adabay_prior`
object passed to `build_cache()`.
