---
title: "Overview of adabay"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview of adabay}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
set.seed(1L)
```

# Motivation

Bayesian group sequential designs (GSDs) deliver the efficiency and ethical
advantages of frequentist GSDs while supporting clinically interpretable
probability statements and the principled incorporation of external evidence.
Their adoption in confirmatory trials has been hindered by the computational
cost of evaluating frequentist operating characteristics at the design stage,
which traditionally requires Markov chain Monte Carlo or another approximate
posterior inference method inside a Monte Carlo trial-simulation loop.

`adabay` implements a *semi-simulation* framework: trial data paths are
generated by Monte Carlo simulation, while per-look posteriors and posterior
tail probabilities are computed analytically or by low-dimensional
deterministic quadrature rather than by MCMC. Conjugacy-preserving updates
are made possible by approximating any user-specified prior by a finite
mixture of conjugate components.

# Architecture at a glance

A typical session builds three or four S3 objects through small constructor
calls and then passes them to the single polymorphic entry point
`evaluate_design()`:

- `set_design()`, `set_prior()`, `set_decision()` and (optionally)
  `set_accrual()` populate the design specification.
- `evaluate_design(design, ...)` (fused path) runs the trial-simulation loop
  for a single design and returns a `adabay_oc` carrying the operating
  characteristics, all in one call.
- `build_cache()` precomputes per-look posterior tail probabilities once;
  `evaluate_design(cache, decision)` (cached path) and `calibrate_design()`
  then score any candidate threshold combination from the same cache in
  milliseconds. The two paths share the `adabay_oc` return contract.
- `fit_mixture()` is a side tributary that approximates a non-conjugate
  prior by a finite mixture of conjugate components, returning a
  `adabay_prior_arm` consumable by `set_prior()`.

The package supports four endpoint types --- continuous, binary, count and
time-to-event --- through their natural conjugate pairs (normal-normal or
normal-inverse-gamma; beta-binomial; gamma-Poisson; gamma-exponential).

# A minimal end-to-end session

We re-implement the ADRENAL trial (Venkatesh et al., 2018) as a five-look
Bayesian GSD on the binary risk-difference scale.

```{r mwe, message = FALSE}
library(adabay)

des <- set_design(
  endpoint     = "binary",
  n_per_look   = c(760, 1520, 2280, 3040, 3800),
  effect_scale = "risk_difference",
  alternative  = "less"
)

pri <- set_prior(
  endpoint = "binary",
  arms     = list(c = list(family = "beta", a = 1, b = 1),
                  t = list(family = "beta", a = 1, b = 1))
)

dec <- set_decision(
  efficacy = list(list(threshold_effect = 0, threshold_prob = 0.99)),
  futility = list(list(threshold_effect = 0, threshold_prob = 0.90)),
  futility_binding = FALSE
)

oc_h1 <- evaluate_design(
  des, pri, dec,
  effect   = list(theta_c = 0.33, theta_t = 0.28),
  n_trials = 2000, cores = 1, seed = 1L
)
oc_h1
```

The returned `adabay_oc` carries the per-look stopping probabilities
(`eff_prob_at_look` / `fut_prob_at_look`, printed as "Stage efficacy" /
"Stage futility"), the expected sample size, and two top-level rates that
are easy to conflate: `$alpha` ("Type I error") and `$eff_prob` ("Efficacy
probability"). They coincide under *binding* futility, but this example
sets `futility_binding = FALSE` (non-binding), where they differ:

- **`$eff_prob`** is the fraction of the *actually simulated* trials that
  crossed efficacy. The simulator always conducts trials under binding
  semantics --- it really does stop the instant either boundary is
  crossed, regardless of `futility_binding` --- so this is real trial
  conduct: sum the printed "Stage efficacy" row and you get `$eff_prob`
  exactly (`0.1970+0.2525+0.1885+0.1515+0.0875 = 0.8770` above).
- **`$alpha`** answers a different, deliberately hypothetical question
  under non-binding futility: what fraction of trials *would* have
  crossed efficacy at some look if the futility rule had been ignored
  entirely and the trial always allowed to continue? This is the standard
  regulatory convention for type I error under a non-binding boundary,
  since a sponsor may always choose to override an advisory futility
  signal and keep testing for efficacy.

`$alpha >= $eff_prob` under non-binding: trials that stop for futility at
an interim look in the real (binding) simulation, but would have gone on
to cross efficacy at a later look had futility never been checked, count
toward `$alpha` but not `$eff_prob`. Set `futility_binding = TRUE` and the
two become identical by construction. `$fut_prob` ("Futility
probability"), by contrast, is unaffected by `futility_binding` --- it is
always just the sum of the "Stage futility" row. See `?evaluate_design`
("Binding vs non-binding futility") for the exact formulas. The
`print()`, `summary()` and `plot()` methods give compact textual and
graphical summaries.

```{r plot-oc, fig.alt = "Per-look stopping probabilities for the ADRENAL re-design under H1."}
if (requireNamespace("ggplot2", quietly = TRUE)) plot(oc_h1)
```

# What the other vignettes cover

Two further vignettes drill into the two main workflows:

- The **evaluation** vignette shows how to evaluate the operating
  characteristics of a fixed design via `evaluate_design()`, with worked
  examples under both a conjugate prior (Beta--Binomial) and a
  non-conjugate prior (logit-normal approximated by a beta mixture from
  `fit_mixture()`).
- The **calibration** vignette shows the design-to-calibration workflow
  through the caching API: `build_cache()` runs the simulation pass once,
  `calibrate_design()` sweeps a grid of posterior-probability thresholds
  against type I error and power targets, and `evaluate_design()` recovers
  stage-specific schedules from the same cache. The same workflow is
  demonstrated under both a conjugate and a non-conjugate prior.

# Cross-`cores` bit-identical reproducibility

By default, `evaluate_design()` and `build_cache()` parallelise the
simulator across `cores` workers, each using its own L'Ecuyer-CMRG
sub-stream. This makes runs at any fixed `(seed, cores)` pair bit-
identical, but the operating characteristics produced at, say, `cores = 1`
and `cores = 8` will differ within Monte Carlo error because the trial-
level draws live on different sub-streams.

When **regulatory-grade reproducibility across hardware configurations**
is required, pass `cross_core_reproducible = TRUE` to `evaluate_design()`
(fused path) and/or `build_cache()`:

```r
oc <- evaluate_design(des, prior = pri, decision = dec,
                      effect = eff,
                      n_trials = 1e5, cores = 8, seed = 1L,
                      cross_core_reproducible = TRUE)
```

The simulator pass is then forced to run in a single driver process
regardless of `cores`, so the trial-level RNG draws coincide with the
`cores = 1` stream at every worker count. The posterior and aggregation
passes are already core-independent. The cost is the loss of simulator
parallelism, which is a small fraction of total wall-clock for binary,
count and tte, and a moderate fraction for continuous. Pair the flag
on `build_cache()` and any downstream `evaluate_design()` /
`calibrate_design()` call when reproducing a cached grid sweep across
hardware.

# Where to look next

- `?adabay` for the high-level package help page.
- `?set_design`, `?set_prior`, `?set_decision`, `?set_accrual` for the
  constructors.
- `?evaluate_design`, `?build_cache`, `?calibrate_design` for the entry
  points.
- `?fit_mixture` for the non-conjugate-prior workflow.
- The companion methodology paper for the semi-simulation framework and
  the per-endpoint posterior derivations.
